Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-09-26 paper/letter.tex
letter diff
submitted version

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Lyman-$\alpha$ Line Sketches a Supermassive Dwarf Galaxy}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ line emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of \sigmaclump with outflows of
\inftyclump) consistent with a dynamical mass of at
least a billion solar masses, $6$ times larger than its baryonic mass.   
We argue that a possible explanation for this excess of
dynamical mass is the presence of a supermassive black hole at the
center of \tol. 
This work demonstrates the importance of considering multiphase
physics and rotation among the possible conditions shaping the
\lya\ spectra of the first galaxies.  
Additionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black
holes.  
}

@@

In this letter we show how the \tol's \lya\ profile can be explained
either by rotation\cite{GaravitoCamargo2014}or the recently developed
class of more complex 
multiphase models that predict a wider variety of spectra
including, single, double and triply peaked spectra \cite{Gronke2016}.  
Figure 1. summarizes our findings.
Dots represent the observational data for \tol\ with the
overplot from our best fits from the analytical solution for a
rotating homogeneous gas sphere (thin line) and the multiphase
model (thick line). 
This is the first time that these models have been introduced with
success to explain an observed \lya\ profile.

@@

Assuming that the clumps are located in a spherical region of radius
$r_s=2.25$ kpc (corresponding to \tol's estimated 3D half-light
radius), 
this corresponds to dynamical masses of $M_{\rm dyn} =
3.2_{-1.0}^{+1.6}\times 10^{10} M_{\odot}$ and $M_{\rm
  dyn}=2.31\pm0.04 \times 10^{9}$ $M_{\odot}$ for the
rotation and multiphase models, respectively,

\tol's stellar mass is  $M_{\star} = 1.45\pm0.45\times 10^{8}
M_{\odot}$   \cite{2014PASP..126.1079M} and its total neutral HI mass is $M_{\rm HI}<2.65\times 10^{8}$ M$_{\odot}$
\cite{pustilnikmartin07}; the dynamical mass is at least 6 to 80 times
the baryonic mass, depending if one considers the multiphase or
rotation estimate.

We lean towards the lower dynamical mass estimate from the multiphase
model as it seems easier to reconcile with the following two
astrophysical mechanisms for its origin.
The first way to explain a dynamical mass of $10^{9} M_{\odot}$
in a sphere of 2 kpc could be having a dark matter halo of at
least $10^{12}$ M$_{\odot}$ in mass \cite{2011ApJ...726..108T}, which
leaves open the question as to why  \tol\ is not more similar to the
Milky Way galaxy as it is hosted by a dark matter halo of similar
mass. 
A second possibility is that \tol\ hosts a supermassive black hole of
$10^{9} M_{\odot}$. This is almost two orders of magnitude higher
than the supermassive black hole found in the compact dwarf galaxy
M60-UCD1 \cite{2014Natur.513..398S}, which has a similar stellar mass
as \tol. This would leave open the question about the formation
process of such a system.

@@

A new observational test is needed to clarify the physical nature of
\tol. 
We suggest that integral field unit measurements spatially
resolving its spatial extent are up to the task. 
\tol\ spans a region of $4$ arcseconds,
an instrument such as the Multi Unit Spectroscopic Explorer
\cite{2014Msngr.157...13B} with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a coarse mapping of different
ionization lines to infer a kinematic map.
Another observational test includes the measurement of the
\lya\ ionizing continuum escape fraction.
In the rotational model this fraction should be zero, while
the multiphase model predicts that averaging over all sightlines
it should be around $0.5^{+1.0}_{-0.4}$\%, with the possibility of strong
variations depending on viewing angle.

@@

The observed flux for the Lyman alpha line is $\sim
8.1\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \cite{Thuan97}
and a Equivalent Width of $70$\AA and its H$\beta$ flux is 
$1.62\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \AA${-1}$
\cite{Izotov04} which gives a Ly$\alpha$/H$\beta$ flux ratio of
4.9$\pm$0.1. The Ly-$\alpha$ flux values correspond to luminosities of
$L_{Ly\alpha}=2.2\times 10^{42}$ erg s$^{-1}$ over a $20$\AA
bandwidth, which in turns translates  into a star formation rate of
$2.0$ M$_{\odot}$ yr$^{-1}$ using a standard conversion factor between
luminosity and star formation rate of $9.1\times 10^{-43}$
$L_{Ly\alpha}$ M$_{\odot}$ yr$^{-1}$. 
The absolute magnitude in the $V$ band translates into a luminosity of
$8.9\times 10^{8}$ L$_{\odot}$.
% using http://tomdwelly.com/tools_fluxtolum.php
Comparing this ratio with the theoretical expectation from case B
recombination of $23.3$ \cite{Hummer1987} one can estimate an escape
fraction of $20$\% for Ly$\alpha$ radiation.
The bolometric UV luminosity is $9.43\pm1.94 \times 10^{8}$
L$_{\odot}$ as measured by GALEX.

@@

The rotation model corresponds to the work presented in
\cite{GaravitoCamargo2014} based on the Monte Carlo code
\texttt{CLARA} \cite{CLARA}. 
In that model the Ly-$\alpha$ photons are propagated 
within a spherical and homogeneous cloud of HI gas undergoing solid
body rotation.
The sphere is fully characterized by three parameters: the HI optical
depth $\tau$ measured from the center to its surface, the HI
temperature $T$, and the linear surface velocity $V_{\rm max}$.  
Photons are emitted at their natural frequency from the center of the
sphere. 
Including the effect of dust only changes the overall line
normalization but not its shape.  
The results we report in the main body of the paper do not include any
dust model.
In this letter we use an analytical solution that captures the most
important effects of rotation onto the \lya\ line.

@@

\cite{GaravitoCamargo2014} derive a concise analytical description for
those features. 
This description takes into account how different parts of the
sphere's surface shift in frequency the \lya\ photons. 
Different shifts in frequency come from different values for the projected
velocity along the LOS. 
Using the analytical solution for the \lya\ spectra of a static sphere
plus the right frequency shifts  computed from geometrical
considerations, one is able to compute an analytical solution for the
rotating sphere that reproduces the main features found using the full
numerical simulation.

@@

\begin{figure}
\begin{center}
\includegraphics[width=1.0\textwidth]{emcee_results.pdf}
\caption{{\bf Results from the Markov Chain Monte Carlo computation for
    the rotation model.} The dotted vertical lines in the outer histograms 
	represent the percentiles 16\%, 50\% and 84\%. \label{emceeresults}} 
\end{center}
\end{figure}

%In the rotation model assuming reasonable bounds for the number
%density of neutral Hydrogen atoms (
%$0.01<n/\mathrm{atoms\ cm^{-3}}<0.1$) and using $\tau=10^7$ the radius
%of the emission region can be bracketed to be in $0.55 < r_s/\mathrm{kpc}< 5.5$.

@@

From this test we found that only three parameters influence the
$\chi^2$: the clump outflow velocity $v_{\infty,\rm{cl}}$ (p-value 
$10^{-18}$), the clump velocity dispersion $\sigma_{\rm cl}$ (p-value
$10^{-4}$) and the probability that the \lya\ emission comes from the
clumps $P_{\rm cl}$ (p-value $10^{-4}$).

@@

%From the multiphase model we obtained $v=$\sigmaclump
%and $r=5$ kpc. This corresponds to a dynamical mass of $M_{\rm
%  dyn}=2.8^{+1.3}_{-1.2}\times 10^{9}$ $M_{\odot}$ which is at least
%$5$ times the HI mass plus the stellar mass estimated from
%observations.

%In the rotation model the size of the spherical region, $r_{s}$, can
%be inferred from the relationship $\tau = \sigma_0 n r_s$, where
%$\sigma_0=5.898\times 10^{-14}$cm$^{-2}$ is the Lyman$\alpha$
%cross section at the  line's center, $n$ is the number density. 
%This expression can be rewritten as
%\begin{equation}
%r_{s} = 0.055 \left(\frac{\tau}{10^7}\right) \left(
%\frac{\mathrm{atoms\ cm^{-3}}}{n}\right) \mathrm{kpc}.
%\end{equation}
%
%Using this result and assuming a bound for the number density of
%$0.01<n/\mathrm{atoms\ cm^{-3}}<0.1$ we have $0.55
%< r_s/\mathrm{kpc}<5.5$.

%This is consistent with the  constraint on the total HI mass and the
%optical depth $\tau_0$ as we show next. The total HI mass can be
%approximated as $M_H = \frac{4\pi}{3} m_H n r_s^3$, where
%$m_H=1.67\times 10^{-24}$ gr is the mass of a single Hydrogen atom.
%This allows us to write the total hydrogen mass as
%\begin{equation}
%M_{H} = 5.70 \times 10^{6} \left(\frac{\tau}{10^{7}}\right)
%\left(\frac{r_s}{\mathrm{kpc}}\right)^2 M_{\odot}
%\end{equation}
%
%The upper limit from the HI mass non-detection imposes $r_{s} <
%6.81$ kpc, which is consistent with the bound we found in the first
%place, as we wanted to show.

%Using those size bounds for the spherical model we find a constrain for
%the dynamical mass of $5.45\times 10^{9}M_{\odot} < M_{\rm dyn} <
%5.45\times 10^{10} M_{\odot}$.

>>>>

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Line Traces of a Supermassive Dwarf Galaxy}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ line emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of \sigmaclump with outflows of
\inftyclump) consistent with a dynamical mass of at
least a billion solar masses, ten times larger than its baryonic mass.   
We argue that a possible explanation for this excess of
dynamical mass is the presence of a supermassive black hole at the
center of \tol. 
This work demonstrates the importance of considering multiphase
physics and rotation among the possible conditions shaping the
\lya\ spectra of the first galaxies.  
Additionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black
holes.  
}

@@

In this letter we show how the \tol's \lya\ profile can be explained
either by rotation \cite{GaravitoCamargo2014} or the recently developed
class of more complex 
multiphase models that predict a wider variety of spectra
including, single, double and triply peaked spectra \cite{Gronke2016}.  
Figure 1. summarizes our findings.
Dots represent the observational data for \tol\ with the
overplot from our best fits from the analytical solution for a
rotating homogeneous gas sphere (thin line) and the multiphase
model (thick line). 
This is the first time that these models have been introduced with
success to explain an observed \lya\ profile.

@@

Assuming that the clumps are located in a spherical region of radius
$r_s=2.25$ kpc (corresponding to \tol's estimated 3D half-luminosity
radius), 
this corresponds to dynamical masses of $M_{\rm dyn} =
3.2_{-1.0}^{+1.6}\times 10^{10} M_{\odot}$ and $M_{\rm
  dyn}=2.31\pm0.04 \times 10^{9}$ $M_{\odot}$ for the
rotation and multiphase models, respectively,

\tol's stellar mass is  $M_{\star} = 1.45\pm0.45\times 10^{8}
M_{\odot}$   \cite{2014PASP..126.1079M} and its total neutral HI mass is $M_{\rm HI}<2.65\times 10^{8}$ M$_{\odot}$
\cite{pustilnikmartin07}; the dynamical mass is at least 12 to 160 times
the baryonic mass, depending if one considers the multiphase or
rotation estimate.

We lean towards the lower dynamical mass estimate from the multiphase
model as it seems easier to reconcile with the following two
astrophysical mechanisms for its origin.
The first way to explain a dynamical mass of $10^{9} M_{\odot}$
in a sphere of 2.25 kpc in radius, could be a dark matter halo of at
least $10^{12}$ M$_{\odot}$ in mass \cite{2011ApJ...726..108T}, which
leaves open the question as to why  \tol\ is not more similar to the
Milky Way galaxy as it would be hosted by a dark matter halo of similar
mass. 
A second possibility is that \tol\ hosts a supermassive black hole of
$10^{9} M_{\odot}$. This is almost two orders of magnitude higher
than the supermassive black hole found in the compact dwarf galaxy
M60-UCD1 \cite{2014Natur.513..398S}, which has a similar stellar mass
as \tol. This would leave open the question about the formation
process of such a system.

@@

A new observational test is needed to clarify the physical nature of
\tol. 
We suggest that integral field unit measurements spatially
resolving its spatial extent are up to the task. 
\tol\ spans a region of $4$ arcseconds,
an instrument such as the Multi Unit Spectroscopic Explorer
\cite{2014Msngr.157...13B} with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a coarse mapping of different
ionization lines to infer a kinematic map.
Another observational test includes the measurement of the
\lya\ ionizing continuum escape fraction.
In the rotational model this fraction should be zero, while
the multiphase model predicts that averaging over all sightlines
it should be around $0.5^{+1.0}_{-0.4}$\%, with the possibility of strong
variations depending on viewing angle \cite{2014MNRAS.444.1095G}.

@@

The observed flux for the Lyman alpha line is $\sim
8.1\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \cite{Thuan97}
and a Equivalent Width of $70$\AA\ and its H$\beta$ flux is 
$1.62\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \AA${-1}$
\cite{Izotov04} which gives a Ly$\alpha$/H$\beta$ flux ratio of
4.9$\pm$0.1. The Ly-$\alpha$ flux values correspond to luminosities of
$L_{Ly\alpha}=2.2\times 10^{42}$ erg s$^{-1}$ over a $20$\AA\ 
bandwidth, which in turns translates  into a star formation rate of
$2.0$ M$_{\odot}$ yr$^{-1}$ using a standard conversion factor between
luminosity and star formation rate of $9.1\times 10^{-43}$
$L_{Ly\alpha}$ M$_{\odot}$ yr$^{-1}$. 
The absolute magnitude in the $V$ band translates into a luminosity of
$8.9\times 10^{8}$ L$_{\odot}$.
% using http://tomdwelly.com/tools_fluxtolum.php
Comparing this ratio with the theoretical expectation from case B
recombination of $23.3$ \cite{Hummer1987} one can estimate an escape
fraction of $20$\% for Ly$\alpha$ radiation.
The bolometric UV luminosity is $9.43\pm1.94 \times 10^{8}$
L$_{\odot}$ as measured by GALEX.

@@

The rotation model corresponds to the work presented in
\cite{GaravitoCamargo2014} based on the Monte Carlo code
\texttt{CLARA} \cite{CLARA}. 
In that model the Ly-$\alpha$ photons are propagated 
within a spherical and homogeneous cloud of HI gas undergoing solid
body rotation.
The sphere is fully characterized by three parameters: the HI line's
center optical  depth $\tau$ measured from the center to its surface, the HI
temperature $T$, and the linear surface velocity $V_{\rm max}$.  
Photons are emitted at their natural frequency from the center of the
sphere. 
Including the effect of dust only changes the overall line
normalization but not its shape.  
The results we report in the main body of the paper do not include any
dust model.
In this letter we use an analytical solution that captures the most
important effects of rotation onto the \lya\ line.

@@

\cite{GaravitoCamargo2014} derived a concise analytical description for
those features. 
This description takes into account how different parts of the
sphere's surface shift in frequency the \lya\ photons. 
Different shifts in frequency come from different values for the projected
velocity along the LOS. 
Using the analytical solution for the \lya\ spectra of a static sphere
plus the right frequency shifts  computed from geometrical
considerations, one is able to compute an analytical solution for the
rotating sphere that reproduces the main features found using the full
numerical simulation.

@@

\begin{figure}
\begin{center}
\includegraphics[width=1.0\textwidth]{emcee_results.pdf}
\caption{{\bf Results from the Markov Chain Monte Carlo computation for
    the rotation model.} The dotted vertical lines in the outer histograms 
	represent the 16th, 50th and 84th percentiles. \label{emceeresults}} 
\end{center}
\end{figure}

@@

From this test we found that only three parameters influence the
$\chi^2$: the clump outflow velocity $v_{\infty,\rm{cl}}$ (p-value 
$10^{-18}$), the clump velocity dispersion $\sigma_{\rm cl}$ (p-value
$10^{-4}$) and the probability that the \lya\ emission comes from the
clumps $P_{\rm cl}$ (p-value $10^{-4}$). This does not mean the other
parameters do not influence the resulting spectra at all; it means
that they cannot be constrained from \tol's observations.

@@
